Using the Design Thinking Empathy Phase as a Facilitator in Privacy Requirements Elicitation

### 2020-02-19 paperRE2020.tex
Updates from Overleaf

\begin{abstract}
%Este artigo apresenta uma contextualização do uso da Phase de Empatia no processo de elicitação de requisitos de privacidade. Nós realizamos uma revisão de literatura para identificar o uso do da Phase de Empatia do Design Thinking na elicitação de requisitos de privacidade. Além disso, nós realizamos um survey com 63 profissionais da indústria para entender como é realizado a elicitação desses requisitos e se estes profissionais utilizam a Phase de Empatia. Nós identificamos que 73.9\% dos desenvolvedores utilizam a fase de empatia para elicitar os requisitos de software. Além disso, mais de 62\% dos profissionais da indústria não possuem conhecimento em relação aos privacy requirements, bem como em relação as ferramentas de Empatia para a requirements elicitation. Por fim, nós propomos um modelo de empatia que poderá ser utilizado pelos desenvolvedores de software no processo de elicitação dos requisitos de privacidade.%

This article presents a contextualization of the use of the Empathy Phase in the process of privacy requirements elicitation. We conducted a literature review to identify the use of the Design Thinking Empathy Phase in privacy requirements elicitation. In addition, we conducted a survey with 68 industry practitioners to understand how these requirements are elicited and whether these practitioners use the Empathy Phase. We found that 73.9\% of developers use the empathy phase to software requirements elicitation. In addition, more than 61\% of industry practitioners are unaware of privacy requirements, as well as empathy tools for requirements elicitation. Finally, we propose an empathy model that can be used by software developers in the process of privacy requirements elicitation.

@@

Developing software that guarantee the protection of personal data is a challenge for software developers, becoming more complex when they involve distributed systems, which can have different ways of collecting data from users, as it happens in most commercial applications and or governmental. In addition, data protection is essential to obtain users' trust and participation in the data collection process. To meet this need for data protection, it is necessary to adopt and use new techniques and tools to requirements elicitation and analyze and make personal data available. One way to facilitate the process of privacy requirements elicitation may be to use the empathy map \cite{DBLP:conf/sbes/FerreiraCB15} and Design Thinking \cite{DBLP:books/lib/brown2009change}, \cite{DBLP:conf/hci/CanedoC18}.

@@

The aim of this research is to identify how the Design Thinking Empathy phase can be used as a facilitating tool in privacy requirements elicitation. Altogether, in this work we answer questions related to (a) the use of the Design Thinking empathy phase in privacy requirements elicitation and (b) the tools that can be used in the empathy phase to elicit the privacy requirements of a company clearer and in a more effective way. Accordingly, we present the following contributions:

\begin{itemize}
    %\item \textcolor{red}{62.9\% dos desenvolvedores da indústria não possuem conhecimento sobre privacy requirements. 27.4\% afirmaram que conhecem sobre privacy requirements e 9.7\% conhecem e trabalham atualmente com a privacy requirements elicitation};
    \item 61.8\% of industry developers are unaware of privacy requirements. 26.5\% said they know about privacy requirements and 11.8\% know and currently work with privacy requirements elicitation;
    %\item  \textcolor{red}{A fase do Design Thinking mais impactante para a privacy requirements elicitation, de acordo com os desenvolvedores de software, é a fase de Emphatia com 65.4\%};
    \item  The most impactful phase of Design Thinking for privacy requirements elicitation, according to software developers, is the Empathy phase with 38.0\%;
     %\item \textcolor{red}{Os desenvolvedores acreditam que a fase de empatia pode motivar os participantes no processo de privacy requirements elicitation, embora a maioria deles não conheçam as ferramentas existentes.}
    \item The developers believe that the empathy phase can motivate participants in the privacy requirements elicitation process, although most of them are unaware of the existing tools.
\end{itemize}

@@

This article is organized as follows. In the section \ref{back} a contextualization is presented about the main concepts related to this research, as well as related works. Section \ref{method} presents the methodology used to conduct this work. Section \ref{empathy} presents the results of the survey carried out to understand the perception of software developers of the use of Empathy phase of Design Thinking in privacy requirements elicitation. The threats to validate this research are presented in Section \ref{thread}. Finally, the conclusions and future work are presented in Section \ref{conclusion}.

@@

Design Thinking considers that the most important task when starting a software development project is to understand the needs of users, investigating what are their needs and what impact the software would cause with its use. All participants in the development process seek to understand the user's universe in order to obtain the necessary information to requirements elicitation and users' perceptions of them \cite{DBLP:books/lib/brown2009change}. Thus, the main idea of Design Thinking is how designers progress with the development process, with creative ideas for the proposed solutions, in order to discover and offer new opportunities \cite{DBLP:conf/hci/AdikariMC13}. Design Thinking disseminates user-centered innovation, so that there is collaboration, interaction and practical approaches to find the most appropriate ideas and, consequently, coherent solutions to users' problems \cite{DBLP:books/lib/brown2009change}.

@@

Newman et al. \cite{DBLP:conf/icse/NewmanFSFFW15} presented the DrivingBoard, a six-phase Design Thinking model that uses Participatory Design and Agile Development. Ximenes et al.  \cite{DBLP:conf/hci/XimenesAA15} combined Design Thinking, Lean Startup and Agile Development. The Design Thinking used by the authors was the model proposed by Stanford University's D-School and in this work we will use this model, which has 5 phases, among them the empathy phase \cite{DBLP:conf/hci/XimenesAA15}, as shown in Figure \ref{fig:design_thinking_phases}:

@@

\item Define: This stage is perhaps one of the most challenging, as it involves the interpretation of all the facts acquired in the process of empathy. For a definition of the problem to be reached, it is necessary to process everything that was said and seen in the Empathy phase, which can be a very laborious and time-consuming activity.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=3.4in]{Figures/FigDR.png}
    \caption{Stanford’s d.school Design Thinking and its Phases \cite{DBLP:conf/hci/XimenesAA15},\cite{ DBLP:conf/icse/CarrollR16}}
    \label{fig:design_thinking_phases}
\end{figure}

@@

Levy and Hadar \cite{DBLP:conf/re/LevyH18} reinforce the need for empathy with end users by stating that this is a requirement engineering skill as necessary as technical knowledge and social competence. Empathy is necessary to work with the diversity of users, allowing a better requirements elicitation. According to Levy and Hadar \cite{DBLP:conf/re/LevyH18}, empathy is the first step of Design Thinking, and it has aroused interest among software development organizations for leveraging design and innovation processes and for enabling better experience for the end user. Design Thinking puts the user in a prominent role, and emphasizes building empathy with users, observing their behavior and drawing conclusions about what users want and need. This work presents preliminary results of a collaboration between design students and engineering students at Shenkar College of Engineering, Design and art. The findings show how engineering students, when dealing with a human being, practiced empathy and used emotional language in Requirement Elicitation for their proposed solutions. The multidisciplinary learning experience of engineering and design students gained from using Design Thinking can promote empathy and other skills needed in modern digital culture, due to the confluence of technology, knowledge and culture.

@@

In the context of Empathy versus Design Thinking, Gasparini \cite{Gasparini} approached Empathy by exploring two main aspects: emotional empathy and cognitive empathy. Emotional or effective empathy refers to an emotional response shared by two or more individuals, for example, feelings of fear or excitement. Cognitive empathy is related to the individual's ability to discover what happens in the other's mind to be able to understand the other's point of view \cite{Newkimbell}. Gasparini \cite{Gasparini} conducted a case study to address the intersection between Design Thinking and Empathy, and presents the results of a workshop organized by the Norwegian University Library addressing Open Access services. The methodologies and techniques used were Photoetnography, Live Notes, User journey, Touch points and Service Design Cards. User journey is the representation of all the steps a customer need to perform to achieve the final goal of the service. Each situation where the user is in direct contact with the service provider is called a touch point. Using Service Design Cards one can make the aforementioned journey, where each card is a touch point. Gasparini \cite{Gasparini} concluded that the use of emotional and cognitive empathy in the design process has good results, but that this subject still needs to be researched, aiming to improve the understanding of how these dimensions of empathy can be used to obtain a perception best suited to the user's needs.

@@

Another technique related to empathy is the Empathy Map (EM). It is a technique that assists in the design of business models according to the users' perspectives. The Empathy Map allows a better understanding of a user's environment, behavior, aspirations and concerns \cite{DBLP:conf/hci/FerreiraBC16}. Thus, it is possible to carry out the survey of implicit requirements \cite{DBLP:conf/re/LevyH18}.

@@

Privacy has become a major concern in software development due to data relating to unauthorized exploitation of data, misuse of information stored on systems, social media sites, internet data and disclosure of personal information to third parties \cite{DBLP:conf/wer/2019}. Therefore, when developing new systems, there is a need to use techniques to privacy requirements elicitation. Several methods can be found in the literature, but in this work four will be listed: Privacy-Friendly System Design, PriS and LINDDUN \cite{DBLP:conf/IEEEares/Beckers12} and a method proposed by Islam et al \cite{DBLP:journals/tcc/IslamOKMG18}.

@@

The PriS method contains a formal approach for modeling of goals and their relationship as a graph. This formal method results in a set of affected processes by the privacy goals.

The LINDDUN method is the acronym of the threats linkability, identifiablitiy, non-repudiation, detectability, information disclosure, content unawareness, and policy/consent noncompliance.

@@

The Privacy-Friendly System Design method proposes an analysis of privacy requirements using important expectations of stakeholders and elements of the information flow: data storage, data processing and data transfer, causing privacy concerns for collection and storage in
personal data, unauthorized use of the personal data collected,
unauthorized access to personal data, errors in personal data,
combination of personal data from various sources.

@@

Considering the context of empathy and software development, Akgun et al. \cite{DBLP:journals/iam/AkgunKCD15} studied collective empathy as a property that provides team members with the ability to identify and work with emotions felt/expressed as a result of group interactions. They tested the role of collective empathy in the effectiveness of the project process. The authors studied 122 software development projects, and found that cognition-based trust, formal communication within the team, and the familiarity of team members can influence the collective empathy of the project teams. They found that collective empathy affects the speed of the product marketing team and learning, and results in lower project development costs \cite{DBLP:journals/iam/AkgunKCD15}.

@@

Only one study was identified that researches the empathy phase of Design Thinking as a facilitator in privacy requirements elicitation. Levy and Hadar \cite{DBLP:conf/re/LevyH18} analyzed empathy as a technique of Design thinking, identifying that software developers, in general, are not concerned with privacy requirements. The analysis of these authors shows that the absence of empathy can lead to the neglect of important aspects of privacy when designing software systems. The authors also state that the empathy phase, in the context of Design Thinking, is a necessary component so that requirements engineering can elicit and deal with requirements with a high risk of being ignored. In addition, the use of empathy techniques and tools provided by Design thinking can promote closer communication and the building of empathy skills for software engineers.

@@

However, it is possible to find relevant works that address requirements elicitation and empathy in general. To improve the requirements elicitation and the experience of using the software, the PATHY technique (Personas empATHY) \cite{DBLP:conf/sbes/FerreiraCB15} was created. This technique integrates the guide questions and the Empathy Map structure with the idea of describing users through personas \cite{DBLP:conf/sbes/FerreiraCB15}. The technique is based on six steps: What the user does; what the user feels or thinks; what is the experience with technology; what are the problems to be solved with the solution to be developed; what are the application features and if there are existing solutions.

Bittner and Shoury \cite{DBLP:conf/hicss/BittnerS19} reconhecem que o Empathy Map Method (EMM) na abordagem Design Thinking é uma ferramenta poderosa, mas são necessárias habilidades metodológicas e a experiência de especialistas para orientar a equipe. Os autores identificam a necessidade de um facilitador profissional de EMM, que possua essas habilidades para disponibilizar esse conhecimento às equipes. Segundo os autores, identificar e contratar um facilitador com essas habilidades específicas é caro e o EMM exige muitas habilidades sociais e cognitivas do facilitador. Assim, os autores propõem uma abordagem para tornar amplamente disponíveis técnicas complexas de colaboração para profissionais sem experiência em EMM por meio da documentação do conhecimento sobre facilitação por meio das abordagens de Engenharia de Colaboração (CE) em projetos de processos estruturados. O conhecimento do método pode ser implementado em sistemas de Tecnologia da Informação (TI) pré-configurados, a fim de executar o processo de maneira semi-automática. Bittner and Shoury \cite{DBLP:conf/hicss/BittnerS19} testaram um piloto com 3 participantes e apresentaram resultados promissores.

@@

To answer the RQ.1 and RQ.3 we conducted a survey with several software developers in order to identify if they work with or know about Privacy Requirements. We also investigate whether developers work with Privacy Requirements Elicitation and whether they use any Empathy tools to facilitate their activities. The questionnaire was divided into two parts: the first part we asked questions to know the profile of the participant, such as, what is his experience, workplace, role, stage of software development that works or worked, if the person knows the phases of Design Thinking, if the person knows about Privacy Requirements and if the person uses or used any Empathy tool for privacy requirements elicitation.

@@

In the second stage of the survey, we investigate which phases of Design Thinking are most impacting for Privacy Requirements Elicitation in the perception of software developers, how the Empathy and Design Thinking phase can motivate the software developer to identify and deal in an effective way with privacy requirements and if empathy is a skill of the developer or should be an initiative of the organization, as well as which tools of Empathy, Creativity and Design Thinking are used by developers in the organizations that they work.

@@

We also performed a focus group with five software developers and security analysts from an organization that work with sensitive user information.

@@

To answer the RQ.2 we performed a literature review to identify the tools that are used in the Empathy phase in Privacy Requirements and which of these tools can facilitate Privacy Requirements Elicitation.

\section{RESULTS}

@@

\begin{mq}
%\emph{"Por meio da empatia é fácil entender o ponto de vista do cidadão/usuário do sistema e suas reais necessidades e limitações."}
\emph{"Through empathy it is easy to understand the point of view of the citizen / user of the system and its real needs and limitations."}
\end{mq}

\begin{mq}
%\emph{"Como o desenvolvedor é também um facilitador, a fase de empatia é a que na minha experiência, mais elucida e anima o desenvolvedor a querer resolver os problemas apontados pelos usuários com que ele se relacionou. Vivenciar a realidade do usuário em muito dá propósito ao trabalho do desenvolvedor, além de permitir insumos para que ele pense em melhores soluções na parte de ideação."}
\emph{"As the developer is also a facilitator, the empathy phase is the one that in my experience, most elucidates and encourages the developer to want to solve the problems pointed out by the users with whom he was related. Experiencing the user's reality in a lot gives purpose to the developer work, in addition to allowing inputs so that he can think of better solutions in the ideation part".}
\end{mq}

@@

\begin{mq}
%\emph{"Se o desenvolvedor conseguir obter uma visão abrangente das restrições do usuário (durante a fase de empatia), ele poderá definir de maneira proativa e precisa as preocupações com a privacidade durante o desenvolvimento do software."}
\emph{"If the developer is able to get a comprehensive view of user restrictions (during the empathy phase), he can proactively and accurately define privacy concerns during software development."}
\end{mq}

@@

We also investigated whether developers are aware of privacy requirements. 61.8\% of developers stated that they do not know about privacy requirements, 26.5\% stated that they know but never worked with privacy requirements and 11.8\% of developers know and actively work with privacy requirements in their daily activities, as shown in Figure \ref{fig:knowledgePR}. After this result, we can conclude that many organizations are not concerned with the privacy of their users' data. Developers who work in the construction of the software, have no knowledge regarding the privacy requirements. On the other hand, those responsible for the data, may not be the developers and this can generate a discussion: How is the data privacy process performed in these organizations, if not through the software?

\begin{figure}[!htb]
    \centering
    \includegraphics[width=3.5in]{Figures/RQ14.png}
    %\caption{Conhecimento dos Desenvolvedores em Relação à Privacy Requirements}
    \caption{Developers' Knowledge of Privacy Requirements}
    \label{fig:conhecimentoPR}
\end{figure}

@@

\todo[inline]{aqui temos que descrever as que achamos na literatura -- Eu só consegui 3 algum tem mais?}

@@

A Web-based system referred to as \href{www.touchpointdashboard.com}{Touchpoint Dashboard} can be used for visualization of software user experience. Visual notations are used for unification of team by conversion of information into a customer journey.  This tool can be employed during the empathy phase \cite{kumar2020methods}.

@@

\begin{figure}
    \centering
    \includegraphics[width=3.5in]{Figures/RQ18.png}
    \caption{Tools related to the Empathy phase be used to facilitate the privacy requirements elicitation}
    \label{fig:tool}
\end{figure}

@@

Regarding the Design Thinking phases that are considered most impactful by the developers in the privacy requirements elicitation, 38\% of the developers stated that it is the Empathy phase, 24\% that is the Definition phase, 16\% the Ideation phase, 14\% the prototyping phase and 8\% stated that it is the Test phase, as shown in Figure \ref{fig:impacts}. From this result, we can conclude that the Empathy phase is considered the most important phase for software developers in privacy requirements elicitation. Thus, an empathy model that can help developers in the execution of the requirements elicitation activity may allow a better understanding of the requirements and a quality elicitation with less errors.

\begin{figure}
    \centering
    \includegraphics[width=3.5in]{Figures/RQ15.png}
    \caption{Phases of Design Thinking are most impactful in privacy requirements elicitation}
    \label{fig:impactos}
\end{figure}

@@

In this question we tried to identify the perception of software developers in relation to empathy skills, investigating whether it should be an orientation of the organization or an initiative of the software developer itself. Some answers we got were:

@@

\todo[inline]{aqui descreveremos os resultados com o grupo focal}

>>>>

\begin{abstract}
%Este artigo apresenta uma contextualização do uso da Phase de Empatia no processo de elicitação de requisitos de privacidade. Nós realizamos uma revisão de literatura para identificar o uso do da Phase de Empatia do Design Thinking na elicitação de requisitos de privacidade. Além disso, nós realizamos um survey com 63 profissionais da indústria para entender como é realizado a elicitação desses requisitos e se estes profissionais utilizam a Phase de Empatia. Nós identificamos que 73.9\% dos desenvolvedores utilizam a fase de empatia para elicitar os requisitos de software. Além disso, mais de 62\% dos profissionais da indústria não possuem conhecimento em relação aos privacy requirements, bem como em relação as ferramentas de Empatia para a requirements elicitation. Por fim, nós propomos um modelo de empatia que poderá ser utilizado pelos desenvolvedores de software no processo de elicitação dos requisitos de privacidade. No grupo focal realizado em duas organizações que trabalham com informações sensíveis dos usuários, nós identificamos que não existe um processo formalizado para realizar a privacy requirements elicitation, embora os membros das equipes de desenvolvimento estejam preocupados com a privacidade dos dados e utilizam algumas técnicas e ferramentas para proteger os dados dos usuários.%

This article presents a contextualization of the use of the Empathy phase in the process of privacy requirements elicitation. We conducted a literature review to identify the use of the Design Thinking Empathy phase in privacy requirements elicitation. In addition, we conducted a survey with 68 industry practitioners to understand how these requirements are elicited and whether these practitioners use the Empathy phase. We found that 73.9\% of developers use the Empathy phase to software requirements elicitation. In addition, more than 61\% of industry practitioners are unaware of privacy requirements, as well as Empathy tools for requirements elicitation. In the focus group conducted at two organizations that work with sensitive user information, we identified that there is no formalized process for conducting privacy requirements elicitation, although members of the development teams are concerned with data privacy and use some techniques and tools to protect user data.

%Finally, we propose an empathy model that can be used by software developers in the process of privacy requirements elicitation.

@@

Developing software that guarantee the protection of personal data is a challenge for software developers, becoming more complex when they involve distributed systems, which can have different ways of collecting data from users, as it happens in most commercial applications and or governmental. In addition, data protection is essential to obtain users' trust and participation in the data collection process \cite{DBLP:journals/access/BinjubeirAISK20}. To meet this need for data protection, it is necessary to adopt and use new techniques and tools to requirements elicitation and analyze and make personal data available. One way to facilitate the process of privacy requirements elicitation may be to use the Empathy Map \cite{DBLP:conf/sbes/FerreiraCB15} and Design Thinking \cite{DBLP:books/lib/brown2009change}, \cite{DBLP:conf/hci/CanedoC18}.

@@

The aim of this research is to identify how the Design Thinking Empathy phase can be used as a facilitating tool in privacy requirements elicitation. Altogether, in this work we answer questions related to (a) the use of the Design Thinking Empathy phase in privacy requirements elicitation and (b) the tools that can be used in the Empathy phase to elicit the privacy requirements of a company clearer and in a more effective way. Accordingly, we present the following contributions:

\begin{itemize}
    %\item \textcolor{red}{62.9\% dos desenvolvedores da indústria não possuem conhecimento sobre privacy requirements. 27.4\% afirmaram que conhecem sobre privacy requirements e 9.7\% conhecem e trabalham atualmente com a privacy requirements elicitation};
    \item 61.8\% of industry developers are unaware of privacy requirements. 26.5\% said they know about privacy requirements and 11.8\% know and currently work with privacy requirements elicitation;
    %\item  \textcolor{red}{A fase do Design Thinking mais impactante para a privacy requirements elicitation, de acordo com os desenvolvedores de software, é a fase de Emphatia com 65.4\%};
    \item  The most impactful phase of Design Thinking for privacy requirements elicitation, according to software developers, is the Empathy phase with 38.0\%;
     %\item \textcolor{red}{Os desenvolvedores acreditam que a fase de empatia pode motivar os participantes no processo de privacy requirements elicitation, embora a maioria deles não conheçam as ferramentas existentes.}
    \item The developers believe that the Empathy phase can motivate participants in the privacy requirements elicitation process, although most of them are unaware of the existing tools.
\end{itemize}

@@

This article is organized as follows. In the section \ref{back} a contextualization is presented about the main concepts related to this research, as well as related works. Section \ref{method} presents the methodology used to conduct this work. Section \ref{result} presents the results of the survey carried out to understand the perception of software developers of the use of Empathy phase of Design Thinking in privacy requirements elicitation. The threats to validate this research are presented in Section \ref{thread}. Finally, the conclusions and future work are presented in Section \ref{conclusion}.

@@

Design Thinking considers that the most important task when starting a software development project is to understand the needs of users, investigating what are their needs and what impact the software would cause with its use \cite{DBLP:conf/icexss/TorresM20}. All participants in the development process seek to understand the user's universe in order to obtain the necessary information to requirements elicitation and users' perceptions of them \cite{DBLP:books/lib/brown2009change}. Thus, the main idea of Design Thinking is how designers progress with the development process, with creative ideas for the proposed solutions, in order to discover and offer new opportunities \cite{DBLP:conf/hci/AdikariMC13}. Design Thinking disseminates user-centered innovation, so that there is collaboration, interaction and practical approaches to find the most appropriate ideas and, consequently, coherent solutions to users' problems \cite{DBLP:books/lib/brown2009change}, \cite{DBLP:journals/ited/SeidelMF20}.

@@

Newman et al. \cite{DBLP:conf/icse/NewmanFSFFW15} presented the DrivingBoard, a six-phase Design Thinking model that uses Participatory Design and Agile Development. Ximenes et al.  \cite{DBLP:conf/hci/XimenesAA15} combined Design Thinking, Lean Startup and Agile Development. The Design Thinking used by the authors was the model proposed by Stanford University's D-School and in this work we will use this model, which has five phases, among them the Empathy phase \cite{DBLP:conf/hci/XimenesAA15}, \cite{\cite{DBLP:journals/ijssmet/Lichtenthaler20}} as shown in Figure \ref{fig:design_thinking_phases}:

@@

\item Define: This stage is perhaps one of the most challenging, as it involves the interpretation of all the facts acquired in the process of Empathy. For a definition of the problem to be reached, it is necessary to process everything that was said and seen in the Empathy phase, which can be a very laborious and time-consuming activity.

@@

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=5.5in]{Figures/FigDR.png}
    \caption{Stanford’s d.school Design Thinking and its Phases \cite{DBLP:conf/hci/XimenesAA15},\cite{ DBLP:conf/icse/CarrollR16}}
    \label{fig:design_thinking_phases}
\end{figure*}

@@

Levy and Hadar \cite{DBLP:conf/re/LevyH18} reinforce the need for Empathy with end users by stating that this is a requirement engineering skill as necessary as technical knowledge and social competence. Empathy is necessary to work with the diversity of users, allowing a better requirements elicitation. According to Levy and Hadar \cite{DBLP:conf/re/LevyH18}, Empathy is the first step of Design Thinking, and it has aroused interest among software development organizations for leveraging design and innovation processes and for enabling better experience for the end user.

Design Thinking puts the user in a prominent role, and emphasizes building Empathy with users, observing their behavior and drawing conclusions about what users want and need \cite{DBLP:conf/re/LevyH18}. This work presents preliminary results of a collaboration between design students and engineering students at Shenkar College of Engineering, Design and art. The findings show how engineering students, when dealing with a human being, practiced Empathy and used emotional language in Requirement Elicitation for their proposed solutions. The multidisciplinary learning experience of engineering and design students gained from using Design Thinking can promote Empathy and other skills needed in modern digital culture, due to the confluence of technology, knowledge and culture.

@@

In the context of Empathy versus Design Thinking, Gasparini \cite{Gasparini} approached Empathy by exploring two main aspects: Emotional Empathy and Cognitive Empathy. Emotional or effective empathy refers to an emotional response shared by two or more individuals, for example, feelings of fear or excitement. Cognitive Empathy is related to the individual's ability to discover what happens in the other's mind to be able to understand the other's point of view \cite{Newkimbell}.

Gasparini \cite{Gasparini} conducted a case study to address the intersection between Design Thinking and Empathy, and presents the results of a workshop organized by the Norwegian University Library addressing Open Access services. The methodologies and techniques used were Photoetnography, Live Notes, User journey, Touch points and Service Design Cards. User journey is the representation of all the steps a customer need to perform to achieve the final goal of the service. Each situation where the user is in direct contact with the service provider is called a touch point. Using Service Design Cards one can make the aforementioned journey, where each card is a touch point. Gasparini \cite{Gasparini} concluded that the use of emotional and cognitive Empathy in the design process has good results, but that this subject still needs to be researched, aiming to improve the understanding of how these dimensions of Empathy can be used to obtain a perception best suited to the user's needs.

@@

Another technique related to Empathy is the Empathy Map (EM). It is a technique that assists in the design of business models according to the users' perspectives. The Empathy Map allows a better understanding of a user's environment, behavior, aspirations and concerns \cite{DBLP:conf/hci/FerreiraBC16}, \cite{DBLP:conf/hicss/BittnerS19}. Thus, it is possible to carry out the survey of implicit requirements \cite{DBLP:conf/re/LevyH18}.

@@

Privacy has become a major concern in software development due to data relating to unauthorized exploitation of data, misuse of information stored on systems, social media sites, internet data and disclosure of personal information to third parties \cite{DBLP:conf/wer/2019}. Therefore, when developing new systems, there is a need to use techniques to privacy requirements elicitation. Several methods can be found in the literature, but in this work four will be listed: Privacy-Friendly System Design \cite{DBLP:journals/access/CaizaMGAY19}, PriS and LINDDUN \cite{DBLP:conf/IEEEares/Beckers12} and a method proposed by Islam et al \cite{DBLP:journals/tcc/IslamOKMG18}.

@@

The PriS method contains a formal approach for modeling of goals and their relationship as a graph. This formal method results in a set of affected processes by the privacy goals \cite{DBLP:journals/re/KalloniatisKG08}, \cite{DBLP:conf/IEEEares/Beckers12}.

The LINDDUN method is the acronym of the threats linkability, identifiablitiy, non-repudiation, detectability, information disclosure, content unawareness, and policy/consent noncompliance \cite{DBLP:conf/IEEEares/Beckers12}.

@@

The Privacy-Friendly System Design method proposes an analysis of privacy requirements using important expectations of stakeholders and elements of the information flow: data storage, data processing and data transfer, causing privacy concerns for collection and storage in personal data, unauthorized use of the personal data collected, unauthorized access to personal data, errors in personal data, combination of personal data from various sources \cite{DBLP:journals/tse/SpiekermannC09}, \cite{DBLP:journals/access/CaizaMGAY19}.

@@

Considering the context of Empathy and software development, Akgun et al. \cite{DBLP:journals/iam/AkgunKCD15} studied collective Empathy as a property that provides team members with the ability to identify and work with emotions felt/expressed as a result of group interactions. They tested the role of collective Empathy in the effectiveness of the project process. The authors studied 122 software development projects, and found that cognition-based trust, formal communication within the team, and the familiarity of team members can influence the collective Empathy of the project teams. They found that collective Empathy affects the speed of the product marketing team and learning, and results in lower project development costs \cite{DBLP:journals/iam/AkgunKCD15}.

@@

Only one study was identified that researches the Empathy phase of Design Thinking as a facilitator in privacy requirements elicitation. Levy and Hadar \cite{DBLP:conf/re/LevyH18} analyzed Empathy as a technique of Design thinking, identifying that software developers, in general, are not concerned with privacy requirements. The analysis of these authors shows that the absence of Empathy can lead to the neglect of important aspects of privacy when designing software systems. The authors also state that the Empathy phase, in the context of Design Thinking, is a necessary component so that requirements engineering can elicit and deal with requirements with a high risk of being ignored. In addition, the use of Empathy techniques and tools provided by Design thinking can promote closer communication and the building of Empathy skills for software engineers.

@@

However, it is possible to find relevant works that address requirements elicitation and Empathy in general. To improve the requirements elicitation and the experience of using the software, the PATHY technique (Personas empATHY) \cite{DBLP:conf/sbes/FerreiraCB15} was created. This technique integrates the guide questions and the Empathy Map structure with the idea of describing users through personas \cite{DBLP:conf/sbes/FerreiraCB15}. The technique is based on six steps: What the user does; what the user feels or thinks; what is the experience with technology; what are the problems to be solved with the solution to be developed; what are the application features and if there are existing solutions.

%Bittner and Shoury \cite{DBLP:conf/hicss/BittnerS19} reconhecem que o Empathy Map Method (EMM) na abordagem Design Thinking é uma ferramenta poderosa, mas são necessárias habilidades metodológicas e a experiência de especialistas para orientar a equipe. Os autores identificam a necessidade de um facilitador profissional de EMM, que possua essas habilidades para disponibilizar esse conhecimento às equipes. Segundo os autores, identificar e contratar um facilitador com essas habilidades específicas é caro e o EMM exige muitas habilidades sociais e cognitivas do facilitador. Assim, os autores propõem uma abordagem para tornar amplamente disponíveis técnicas complexas de colaboração para profissionais sem experiência em EMM por meio da documentação do conhecimento sobre facilitação por meio das abordagens de Engenharia de Colaboração (CE) em projetos de processos estruturados. O conhecimento do método pode ser implementado em sistemas de Tecnologia da Informação (TI) pré-configurados, a fim de executar o processo de maneira semi-automática. Bittner and Shoury \cite{DBLP:conf/hicss/BittnerS19} testaram um piloto com 3 participantes e apresentaram resultados promissores.

@@

To answer the RQ.1 and RQ.3 we conducted a survey with several software developers in order to identify if they work with or know about privacy requirements. We also investigate whether developers work with privacy requirements elicitation and whether they use any Empathy tools to facilitate their activities. The questionnaire was divided into two parts: the first part we asked questions to know the profile of the participant, such as, what is his experience, workplace, role, stage of software development that works or worked, if the person knows the phases of Design Thinking, if the person knows about privacy requirements and if the person uses or used any Empathy tool for privacy requirements elicitation.

@@

In the second stage of the survey, we investigate which phases of Design Thinking are most impacting for privacy requirements elicitation in the perception of software developers, how the Empathy and Design Thinking phase can motivate the software developer to identify and deal in an effective way with privacy requirements and if Empathy is a skill of the developer or should be an initiative of the organization, as well as which tools of Empathy, Creativity and Design Thinking are used by developers in the organizations that they work.

@@

We also performed a focus group \cite{DBLP:conf/isese/KontioLB04} with one practitioner user experience, two security analyst, two software developers and two requirements analyst from an organization that work with sensitive user information.

@@

To answer the RQ.2 we performed a literature review to identify the tools that are used in the Empathy phase in privacy requirements and which of these tools can facilitate privacy requirements elicitation.

\section{RESULTS}
\label{result}

@@

\begin{mq}
%\emph{"Por meio da empatia é fácil entender o ponto de vista do cidadão/usuário do sistema e suas reais necessidades e limitações."}
\emph{"Through Empathy it is easy to understand the point of view of the citizen / user of the system and its real needs and limitations."}
\end{mq}

\begin{mq}
%\emph{"Como o desenvolvedor é também um facilitador, a fase de empatia é a que na minha experiência, mais elucida e anima o desenvolvedor a querer resolver os problemas apontados pelos usuários com que ele se relacionou. Vivenciar a realidade do usuário em muito dá propósito ao trabalho do desenvolvedor, além de permitir insumos para que ele pense em melhores soluções na parte de ideação."}
\emph{"As the developer is also a facilitator, the Empathy phase is the one that in my experience, most elucidates and encourages the developer to want to solve the problems pointed out by the users with whom he was related. Experiencing the user's reality in a lot gives purpose to the developer work, in addition to allowing inputs so that he can think of better solutions in the ideation part".}
\end{mq}

@@

\begin{mq}
%\emph{"Se o desenvolvedor conseguir obter uma visão abrangente das restrições do usuário (durante a fase de empatia), ele poderá definir de maneira proativa e precisa as preocupações com a privacidade durante o desenvolvimento do software."}
\emph{"If the developer is able to get a comprehensive view of user restrictions (during the Empathy phase), he can proactively and accurately define privacy concerns during software development."}
\end{mq}

@@

We also investigated whether developers are aware of privacy requirements. 61.8\% of developers stated that they do not know about privacy requirements, 26.5\% stated that they know but never worked with privacy requirements and 11.8\% of developers know and actively work with privacy requirements in their daily activities, as shown in Figure \ref{fig:know}. %After this result, we can conclude that many organizations are not concerned with the privacy of their users' data. Developers who work in the construction of the software, have no knowledge regarding the privacy requirements. On the other hand, those responsible for the data, may not be the developers and this can generate a discussion: How is the data privacy process performed in these organizations, if not through the software?

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=5.5in]{Figures/RQ14.png}
    %\caption{Conhecimento dos Desenvolvedores em Relação à Privacy Requirements}
    \caption{Developers' Knowledge of Privacy Requirements}
    \label{fig:know}
\end{figure*}

@@

\todo[inline]{aqui temos que descrever as que achamos na literatura -- Eu só consegui 3 -- Angelica você pode fechar essa questão?}

@@

A Web-based system referred to as \href{www.touchpointdashboard.com}{Touchpoint Dashboard} can be used for visualization of software user experience. Visual notations are used for unification of team by conversion of information into a customer journey.  This tool can be employed during the Empathy phase \cite{kumar2020methods}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=5.5in]{Figures/RQ18.png}
    \caption{Tools related to the Empathy phase be used to facilitate the privacy requirements elicitation}
    \label{fig:tool}
\end{figure*}

@@

Regarding the Design Thinking phases that are considered most impactful by the developers in the privacy requirements elicitation, 38\% of the developers stated that it is the Empathy phase, 24\% that is the Definition phase, 16\% the Ideation phase, 14\% the prototyping phase and 8\% stated that it is the Test phase, as shown in Figure \ref{fig:impactos}. %From this result, we can conclude that the Empathy phase is considered the most important phase for software developers in privacy requirements elicitation. Thus, an empathy model that can help developers in the execution of the requirements elicitation activity may allow a better understanding of the requirements and a quality elicitation with less errors.

\begin{figure*}
    \centering
    \includegraphics[width=5.5in]{Figures/RQ15.png}
    \caption{Phases of Design Thinking are most impactful in privacy requirements elicitation}
    \label{fig:impactos}
\end{figure*}

@@

In this question we tried to identify the perception of software developers in relation to Empathy skills, investigating whether it should be an orientation of the organization or an initiative of the software developer itself. Some answers we got were:

@@

%Para complementar as respostas às questões de pesquisa nós realizamos um grupo focal com os desenvolvedores de duas organizações que trabalham com informações sensíveis do usuário com o objetivo de verificar como a privacy requirements elicitation é realizada e identificar a percepção dos desenvolvedores em relação ao uso as ferramentas e técnicas da Empatia.
